---
title: "SCH6245 Clinical Bioinformatics"
subtitle: "Module 1: Reproducible Research Example"
author: "Dr Alyce Russell <br> Postdoctoral Research Fellow <br> Centre for Precision Health, ECU <br> a.russell@ecu.edu.au <br> https://www.twitter.com/nerdrusty "
date: "Created: 3 March 2021 <br> Last Updated: `r format(Sys.Date(), format = '%d %B %Y')`"
output: 
  html_document:
    css: buttons.css
    fig_caption: yes
    highlight: tango
    keep_md: yes
    theme: lumen
    toc: yes
    toc_float:
      collapsed: false
    toc_depth: 2
    self_contained: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

```{r, echo=FALSE}
#Determine the output format of the document
outputFormat = knitr::opts_knit$get("rmarkdown.pandoc.to")

#Figure and Table Caption Numbering, for HTML do it manually
capTabNo = 1; capFigNo = 1;

#Function to add the Table Number
capTab = function(x){
  if(outputFormat == 'html'){
    x = paste0("**", "Table ", capTabNo, ". ", "**", x)
    capTabNo <<- capTabNo + 1
  }; x
}

#Function to add the Figure Number
capFig = function(x){
  if(outputFormat == 'html'){
    x = paste0("**", "Figure ",capFigNo,". ", "**", x)
    capFigNo <<- capFigNo + 1
  }; x
}

```

<div style="margin-bottom:50px;">

<style>
 #TOC {
  background: url("./images/Centre-for-Precision-Health.png");
  background-size: contain;
  padding-top: 100px !important;
  background-repeat: no-repeat;
}
</style>

</div>

<style>
  p.caption {
    font-size: 1.2em;
    }
  caption {
    font-size: 1.2em;
    } 
  body{
    font-size: 12pt;
    }
</style>


<script src="hideOutput.js"></script>   


## Overview

<br> 

This is a walk-through of the first activity for the semester. I purposely gave you little direction to get you started on decision making required in bioinformatics. You may have chosen different examples, which is fine as you will also be doing different projects. Hopefully this will explain a few things that were confusing. 

Make sure you have a go doing this yourself and submit your final output to "*Discussion Board > Week 1 Output*". Also, check each others content. Your peers may know how to do something more efficiently. 

<br> 

***

## Load Dependencies


```{r}
library(tidyverse)
library(readxl)
library(kableExtra)
library(ggpubr)
```

## Data Preparation

```{r}
# first list the sheet names for your info
excel_sheets("./data/surveys_data.xlsx")

# load the data first
survey_data <- read_excel("./data/surveys_data.xlsx", sheet = 1)

# load in plot info
plot_info <- read_excel("./data/surveys_data.xlsx", sheet = 2)

# load in species info
species_info <- read_excel("./data/surveys_data.xlsx", sheet = 3)
```

<br>

It is good practice to always check your data structure first and make sure your variables are in the right format. Then you should clean variables and add value labels. Bioinformatics and statistics is about 80% data preparation and 20% analysis and visualisation. 

<br>

<div class="fold o">

```{r}
# check structure of the data - this function shows variable names (next to `$`), 
# variable format, and the first few lines of code. 
str(survey_data)

# ok so first off I am going to recode sex as numeric.
# to recode sex and set value labels, you can use something like below 
# (though other options exist).
survey_data$sex <- ifelse(survey_data$sex=="M", 0, 
                          ifelse(survey_data$sex=="F", 1, NA))
survey_data$sex <- factor(survey_data$sex,
                          levels=c(0,1),
                          labels=c("Male", "Female"))
```

</div>

<br>

Label plot_id in main dataset using the variables in your `plot_info` data. 

<br>

```{r}
# explore the data structure to determine steps required
str(plot_info)  
head(plot_info)  #prints first few lines of data so you can see what the format is

# label plot_id according to the plot_type info in thhe plot_info dataset
survey_data$plot_id <- factor(survey_data$plot_id,          # this is the variable we want label
                              levels=plot_info$plot_id,     # these are the levels of the factors
                              labels=plot_info$plot_type)   # these are the value labels matching those levels
```

<br>

Label `species_id` in main dataset using the variables in your `species_info` data.

<br>

<div class="fold o">

```{r}
# explore the data structure to determine steps required
str(species_info)

# next merge `genus` and `species` variables into one column.
# NOTE: this last bit adds a space between genus and species
species_info$species_lab <- paste(species_info$genus, species_info$species, " ") 

# check the first 10 rows using this code
species_info[1:10,]  # head(species_info) will give you the first 6 rows

# there is lagging space present, which I want to remove
species_info$species_lab <- trimws(species_info$species_lab, which = "right")
species_info[1:10,]

# label the species id values in survey_data -- same as above
survey_data$species_id <- factor(survey_data$species_id, 
                                 levels=species_info$species_id,
                                 labels=species_info$species_lab)

# add genus and taxa to your data too
survey_data <- merge(survey_data, species_info[,c(2,4,5)], 
                     by.x="species_id", by.y="species_lab")

str(survey_data)
head(survey_data)

# remove labels with missing
survey_data$species_id <- droplevels(survey_data$species_id)
```

</div>

<br>

We noticed some weirdness in the data on campus. This is common and the dataset was not clean. For the purpose of this exercise, let's remove any with missing `sex` or `weight`.

<br>


```{r}
# new dataset without missing data for sex or weight
survey_data$sexmiss <- ifelse(is.na(survey_data$sex), 1, 0)  # creating index number to indicate missing sex
survey_data$weightmiss <- ifelse(is.na(survey_data$weight), 1, 0)
survey_data$anymiss <- ifelse(survey_data$sexmiss==1 | survey_data$weightmiss==1, 1, 0)  # missing either

table(survey_data$sexmiss, survey_data$anymiss)    # check all sexmiss are counted
table(survey_data$weightmiss, survey_data$anymiss) # ditto for weightmiss
table(survey_data$anymiss)  #2604 to be excluded

survey_data2 <- survey_data[!survey_data$anymiss==1, ]    # new data excluding missing by indicator
dim(survey_data)  # 34786    15  (read as row x column)
dim(survey_data2) # 32182    15  (34786 - 2604 = 32182... confirm exclusion worked)
```

<div class="fold o">

```{r}
str(survey_data2)
head(survey_data2)
names(survey_data2)  #prints name and column number -- use to exclude unwanted cols
survey_data2 <- survey_data2[, c(1:12)] 

```

</div>

<br>

Last thing I'll do is create a season variable from the date. Season can affect many things in the animal kingdom as well as in bioinformatics studies. 

<br>

>What are some examples you can think of??

<br>

```{r}
# create season variable with a nested ifelse function -- this is intermediate level but USEFUL to know
survey_data2$season <- ifelse(survey_data2$month==12 | survey_data2$month==1 | survey_data2$month==2, 0,  #if Dec to Feb, code '0', else.. 
                              ifelse(survey_data2$month==3 | survey_data2$month==4 | survey_data2$month==5, 1, 
                                     ifelse(survey_data2$month==6 | survey_data2$month==7 | survey_data2$month==8, 2, 
                                            ifelse(survey_data2$month==9 | survey_data2$month==10 | survey_data2$month==11, 3, NA)))) 
# check
table(survey_data2$season)

# label seasons variable
survey_data2$season <- factor(survey_data2$season,
                              levels=c(0,1,2,3), #telling it these are the levels
                              labels=c("Summer", "Autumn", "Winter", "Spring"))
table(survey_data2$season) # labels have worked

```

<br>

***

## Exploratory Data Analysis (EDA)

<br>

<div class="fold s">

```{r}
# counts
summSeason <- survey_data2 %>%
  group_by(season) %>%
  count() %>%  #count by season
  as.data.frame() #output as data.frame not tibble
summSeason$p <- round(summSeason$n/nrow(survey_data2)*100,1)  #create percentages

summSex <- survey_data2 %>%
  select(season, sex) %>%
  group_by(season) %>%
  count(sex) %>%
  as.data.frame()
summSex$p <- c(round(summSex$n[1:2]/sum(summSex$n[1:2])*100,1),
               round(summSex$n[3:4]/sum(summSex$n[3:4])*100,1),
               round(summSex$n[5:6]/sum(summSex$n[5:6])*100,1),
               round(summSex$n[7:8]/sum(summSex$n[7:8])*100,1))  #create percentages by columns

# means
summMeans <- survey_data2 %>%
  select(season, hindfoot_length, weight) %>%
  group_by(season) %>%
  summarise(across(.cols = everything(), ~mean(., na.rm = TRUE))) %>%
  as.data.frame()

# sd
summSD <- survey_data2 %>%
  select(season, hindfoot_length, weight) %>%
  group_by(season) %>%
  summarise(across(.cols = everything(), ~sd(., na.rm = TRUE))) %>%
  as.data.frame()

# plot by season - do they differ over the year??
summPlots <- survey_data2 %>%
  select(season, plot_id) %>%
  group_by(season) %>%
  count(plot_id) %>%
  as.data.frame()
summPlots$p <- c(round(summPlots$n[1:5]/sum(summPlots$n[1:5])*100,1),
                 round(summPlots$n[6:10]/sum(summPlots$n[6:10])*100,1),
                 round(summPlots$n[11:15]/sum(summPlots$n[11:15])*100,1),
                 round(summPlots$n[16:20]/sum(summPlots$n[16:20])*100,1))
```

<br>

To create summary table using `kableExtra` and my output from `dplyr`, first I create a data frame that prints exactly how I would want it to look (i.e. vars in first column then subsequent columns are some condition).

<br>


```{r}
t1 <- data.frame(vars =   c(paste0("**Overall Count** "," *n (%)*"), 
                            paste0("**Sex** "," *n (%)*"), 
                                   "*Female*", "*Male*", 
                            paste0("**Hindfoot Length** ", "mm", " *mean (SD)*"),
                            paste0("**Weight** ", "g", " *mean (SD)*"),
                            paste0("**Plot Information** ", "*n (%)*"),
                                  "*Spectab Exclusure*", "*Control*", "*Long-term Krat Exclosure*", 
                                  "*Rodent Exclosure*", "*Short-term Krat Exclusure*"),
                 
                 summer = c(paste0(summSeason$n[1], " (", summSeason$p[1], "%)"),
                            "",  # add space for sex title
                            paste0(summSex$n[1:2], " (", summSex$p[1:2], "%)"),
                            paste0(round(summMeans$hindfoot_length[1],2), " (", 
                                   round(summSD$hindfoot_length[1], 2), ")"),
                            paste0(round(summMeans$weight[1], 2), " (", 
                                   round(summSD$weight[1], 2), ")"),
                            "",  # space for Plot Info title
                            paste0(summPlots$n[1:5], " (", summPlots$p[1:5], "%)")),
                 
                 autumn = c(paste0(summSeason$n[2], " (", summSeason$p[2], "%)"), 
                            "",  
                            paste0(summSex$n[3:4], " (", summSex$p[3:4], "%)"),
                            paste0(round(summMeans$hindfoot_length[2],2), " (", 
                                   round(summSD$hindfoot_length[2], 2), ")"),
                            paste0(round(summMeans$weight[2], 2), " (", 
                                   round(summSD$weight[2], 2), ")"),
                            "",  
                            paste0(summPlots$n[6:10], " (", summPlots$p[6:10], "%)")),
                 
                 
                 winter = c(paste0(summSeason$n[3], " (", summSeason$p[3], "%)"), 
                            "",  
                            paste0(summSex$n[5:6], " (", summSex$p[5:6], "%)"),
                            paste0(round(summMeans$hindfoot_length[3],2), " (", 
                                   round(summSD$hindfoot_length[3], 2), ")"),
                            paste0(round(summMeans$weight[3], 2), " (", 
                                   round(summSD$weight[3], 2), ")"),
                            "", 
                            paste0(summPlots$n[11:15], " (", summPlots$p[11:15], "%)")),
                 
                 spring = c(paste0(summSeason$n[4], " (", summSeason$p[4], "%)"), 
                            "",  
                            paste0(summSex$n[7:8], " (", summSex$p[7:8], "%)"),
                            paste0(round(summMeans$hindfoot_length[4],2), " (", 
                                   round(summSD$hindfoot_length[4], 2), ")"),
                            paste0(round(summMeans$weight[4], 2), " (", 
                                   round(summSD$weight[4], 2), ")"),
                            "", 
                            paste0(summPlots$n[16:20], " (", summPlots$p[16:20], "%)")),
                 
                 stringsAsFactors = FALSE)
```

</div>

<br>

Here's what the dataframe looks like before formatting a table. 

<br>

```{r}
t1
```


<br>

You can create amazing tables using `kableExtra`. A good reference can be found [here](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html). These tables are fully customisable, though we'll do a basic summary table.

You can use my [previous workshop](https://rpubs.com/R-LadiesPerth/creating-tables-rmarkdown) as a reference. 

<br>

```{r}
kable(t1,
      caption = capTab("**Summary of study variables by season**"), # adds table caption
      col.names = c("", "Summer", "Autumn", "Winter", "Spring"),   # no title to first column
      align="lcccc", 
      type="") %>%
  column_spec(1, width_min="5cm", border_right = TRUE) %>%
  column_spec(c(2,3,4,5), width_min="3cm") %>%
  row_spec(0, bold=T, color="white", background="#666666") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  add_indent(c(3,4,8:12)) %>%  # indent rows for measures with levels
  footnote(general = c("*% are within groups for all non-missing values*")) %>% # add footnote
  add_header_above(c(" " = 1, "Season" = 4), bold=T, color="white", background="#666666", include_empty = T)
```

<br> 

***

## Data Visualisation

<br>

You can use my [previous workshop](https://rpubs.com/r-ladiesperth/datavisualisationbasics) and the free eBook [ggplot2: Elegant Graphics for Data Analysis](https://ggplot2-book.org/) as references to create some plots. Here are some examples below. 

<br>

<div class="fold s">

```{r}
# remove levels where species_id has 0 records
survey_data2$species_id <- droplevels(survey_data2$species_id) 
survey_data2$genus <- as.factor(survey_data2$genus)

# number samples collected each year
p1 <- survey_data2 %>%
  select(year, genus) %>%
  group_by(year) %>%
  count(genus) %>%
  ggplot( aes(x=year, y=n, colour=genus)) +
  geom_line(size=1.2) +
  theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  xlab("Year") + ylab("Number of Samples Collected") 

# weight vs length by plot_id colour=species_id
p2 <- ggplot(survey_data2, aes(x=hindfoot_length, y=weight, colour=as.factor(genus))) +
  geom_point(na.rm = TRUE, alpha=0.5) + theme_bw() +
  xlab("Hindfoot Length (mm)") + ylab("Weight (g)") + 
  facet_grid(~ plot_id) + 
  scale_colour_discrete(name = "Genus") + theme(legend.position = "none")
  
# length by sex
p3 <- ggplot(survey_data2, aes(x=hindfoot_length, fill=sex)) +
  geom_histogram(aes(y = ..density..), colour="black", bins=50, alpha=0.5) + 
  geom_density(size=1, alpha=0.4) + 
  theme_bw() + facet_grid(sex ~ .) + theme(legend.position = "none") +
  xlab("Hindfoot Length (mm)") + ylab("Density")

p3a <- ggplot(survey_data2, aes(x=as.factor(genus), y=hindfoot_length, colour = as.factor(genus))) +
  geom_boxplot(fill="white", weight=1.5) + 
  geom_point(#fill="white", 
             position="jitter", 
             alpha=0.1) +
  xlab("Genus") + ylab("Hindfoot Length (mm)") +
  theme_bw() + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90)) +
  coord_flip()

p4 <- ggplot(survey_data2, aes(x=weight, fill=sex)) +
  geom_histogram(aes(y = ..density..), colour="black", bins=50, alpha=0.5) + 
  geom_density(size=1, alpha=0.4) + 
  theme_bw() + facet_grid(sex ~ .) + theme(legend.position = "none") +
  xlab("Weight (g)") + ylab("Density")

p4a <- ggplot(survey_data2, aes(x=as.factor(genus), y=weight, colour = as.factor(genus))) +
  geom_boxplot(fill="white", weight=1.5) + 
  geom_point(#fill="white", 
             position="jitter", 
             alpha=0.1) +
  xlab("Genus") + ylab("Weight (g)") +
  theme_bw() + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90)) +
  coord_flip()
```

<br>

```{r, fig.align="center", fig.width=8, fig.height=5, fig.cap =capFig("**Visualisations of Genera. a)** Hindfoot Length vs Weight, **b)** Number of Samples Collected")}
ggarrange(
  p2, p1,
  labels = c("a", "b"),
  nrow = 2#, heights = c(1,1.5)
)
```

<br>

```{r, fig.align="center", fig.width=8, fig.height=5, fig.cap =capFig("**Distributions of Hindfoot Length. a)** By Sex, **b)** By Genus")}
ggarrange(
  p3, p3a,
  labels = c("a", "b")
)
```

<br>

```{r, fig.align="center", fig.width=8, fig.height=5, fig.cap =capFig("**Distributions of Weight. a)** By Sex, **b)** By Genus")}
ggarrange(
  p4, p4a,
  labels = c("a", "b")
)
```

</div>

<br>

***

```{r}
sessionInfo()
```
